---
title: "Santa Barbara urchins"
author: "Halley McVeigh"
date: "11/1/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(janitor)
library(here)
library(kableExtra)
library(lubridate)

```

### Read in Santa Barbara urchins data

```{r}
urchins <- read_csv(here("data", "urchins.csv")) %>% 
  clean_names()
```

## Section 1. Creating exploratory data visualizations

### Data visualization: Faceted histogram 

```{r}
ggplot(data = urchins) +
  geom_histogram(aes(x = size, fill = common_name), alpha = 0.75, binwidth = .5, color = "gray") +
  facet_wrap(~common_name) + # separate histogram plots for each common name
  labs(x = "Size (cm)", y = "Count") + # rename axes
  theme(legend.position = "none") + # remove legend
  scale_fill_manual(values = c("purple", "red")) # fill color for urchin color
```


### Data visualization: Boxplot
```{r}
ggplot(data = urchins) +
  geom_boxplot(aes(x = common_name, y = size)) +
  labs(x = "Common name", y = "Size (cm)")
```

## Descriptive statistics table
A summary table to include **mean, median, standard deviation, and sample size** for purple and red urchin data.

```{r}
urchins_summary <- urchins %>% 
  group_by(common_name) %>% 
  summarize(mean_size = mean(size, na.rm = T), # calculate mean, median, sd, n
            median_size = median(size, na.rm = T),
            standard_deviation = sd(size, na.rm = T),
            sample_size = n())

kbl(urchins_summary, caption = "Urchin Summary Table", # create summary table
    col.names = c("Common name", # rename column names
                  "Mean size (cm)",
                  "Median size (cm)",
                  "Standard deviation",
                  "Sample size")) %>% 
  kable_styling(full_width = FALSE, "bordered")
```


The distribution of the red urchin is...



## Section 2. Exploring purple urchin counts by site and year

```{r}
purple_urchins <- urchins %>%
  filter(common_name == "Purple Urchin") %>% # include only purple urchins
  mutate(date = mdy(date)) %>% # convert the date to class "Date"
           mutate(year = year(date)) %>% # creating a column with year
  mutate(site_full = case_when( # creating a new column with full names of sites
    site %in% c("NAPL") ~ "Naples Reef",
    site %in% c("AQUE") ~ "Arroyo Quemado Reef",
    site %in% c("IVEE") ~ "Isla Vista Reek",
    site %in% c("CARP") ~ "Carpinteria Reef",
    site %in% c("MOHK") ~ "Mohawk Reef"))
  
```


A summary table containing the **total number of purple urchins observed by site and year**

```{r}
purple_urchins_summary <- purple_urchins %>% 
  select("year", "site_full") %>% 
  filter(!is.na(year)) %>% # filter out all rows where year is na
  count(site_full, year) # count() combines group_by and summarize

kbl(purple_urchins_summary, caption = "Purple urchin count by site and year", # create a summary table
    col.names = c("Site", "Year", "Number observed")) %>% 
  kable_styling(full_width = TRUE, "bordered")
```

An exploratory graph of counts of purple urchins recorded at each site for the years included in the study

```{r}
ggplot(data = purple_urchins_summary) + 
  geom_col(aes(x = year, y = n, fill = site_full),
           color = "black", alpha = .75, bins = 30) + # fill each site different color, outline in black
  facet_wrap(~site_full) + # create different bar graphs for each site
  scale_x_continuous(breaks = seq(2000, 2012, by = 4)) + # years on x axis labeled every 4 years
  theme(legend.position = "none") +
  labs(x = "Year", y = "Count observed", title = "Purple urchins of Santa Barbara")
```

