---
title: "Naples Reef (January 2010) purple urchins"
author: "Halley McVeigh"
date: "11/4/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(janitor)
library(lubridate)
library(here)
```

```{r}
urchins <- read_csv(here("data", "urchins.csv")) %>% 
  clean_names()
```

```{r}
naples <- urchins %>%
  filter(common_name == "Purple Urchin") %>% 
  filter(site == "NAPL") %>% 
  mutate(date = mdy(date)) %>% 
  mutate(year = year(date)) %>% 
  mutate(month = month(date)) %>% 
  filter(year == 2010) %>% 
  filter(month == 1)
  
```

### Exploratory histogram
```{r}
ggplot(data = naples) +
  geom_histogram(aes(x = size, fill = treatment), breaks = seq(2,6,.5), alpha = .75) +
  facet_wrap(~treatment) +
  labs(x = "Size (cm)", y = "Count") + # rename axes
  theme(legend.position = "none") + # remove legend
  scale_fill_manual(values = c("purple", "darkgreen")) # fill colors
```
Somewhat normally distributed


### A quantile-quantile plot to explore normality
```{r}
ggplot(data = naples) +
  geom_qq(aes(sample = size)) +
  facet_wrap(~treatment) # separate qq plot by treatment
```
The closer it is to linear, the closer the sample is normally distributed. They both generally follow a linear trend.


### Descriptive statistics
```{r}
naples_summary <- naples %>% 
  group_by(treatment) %>% 
  summarize(mean_size = mean(size, na.rm = TRUE),
            median_size = median(size, na.rm = TRUE),
            sd_size = sd(size, na.rm = TRUE),
            n = n())

naples_summary
```
The sample sizes are over 60, well over 30, the sampling distribution of means will be normal. A t-test is therefore appropriate to compare the means of sizes between treatments of purple urchins observed at the Naples site. 

### Running a two sample t-test 
Create vectors containing observations for each treatment sample using `filter()` and `pull()` to extract the size column as a vector

```{r}
annual_sample <- naples %>% 
  filter(treatment == "ANNUAL") %>% 
  pull(size) # pull size column in annual treatment as a vector

control_sample <- naples %>% 
  filter(treatment == "CONTROL") %>% 
  pull(size) # pull size column in control treatment as a vector

t.test(annual_sample, control_sample)
```

The mean of annual treatment is 3.92 and the mean of the control treatment is 3.73. The null hypothesis that is tested in this two sample t-test is that the means of the two samples are equal, or that the difference in means is equal to zero. In this test there is a 25% chance that samples could have been randomly drawn with means at least as different, if they were taken from populations with the same mean. If 
The explanation for this is that these samples could have been taken from a sa








